Week 8 Accessibility Analysis

```{r}
library(here)
library(usethis)
library(tidyverse)
library(sf)
library(osmdata)
library(dodgr)
library(tmap)
```
```{r}
# read poi data
poi24 <- st_read(here::here("Dataset", "lambeth-POI-2024.gpkg"))
```

```{r}
# read LSOA dataset
lsoa11 <- st_read(here::here("Dataset", "london-LSOA-2011.gpkg"))
```

```{r}
# read borough dataset
borough <- st_read(here::here("Dataset", "London-Boroughs.gpkg"))
```

Examine individual dataframe

```{r}
# inspect individual dataframe
head(poi24)
head(lsoa11)
head(borough)
```

Start by filtering out all POIs where the word `school` features in the `main_category` column

```{r}
# filter school poi data
poi_schools <- poi24 %>% 
  filter(str_detect(main_category, "school"))

# inspect
head(unique(poi_schools$main_category), n=50)
```

```{r}
# remove quotes for easier processing
poi_schools <- poi_schools %>% 
  mutate(main_category = str_replace_all(main_category, "\"", ""))

# filter school poi data
poi_schools <- poi_schools %>% 
  filter(main_category == "elementary_school" |
           main_category == "high_school" |
           main_category == "middle_school" |
           main_category == "private_school" |
           main_category == "public_school" |
           main_category == "school")

# filter school poi data
lambeth <- borough %>% 
  filter(name == "Lambeth")

poi_schools <- poi_schools %>% 
  st_intersection(lambeth) %>% 
  select(1:11)

# inspect
poi_schools
```

We can use the `alternate_category` column to further narrow down our results

```{r}
# filter school poi data
poi_schools <- poi_schools %>% 
  filter(str_detect(alternate_category, "elementary_school") |
           str_detect(alternate_category, "high_school") |
           str_detect(alternate_category, "middle_school") |
           str_detect(alternate_category, "private_school") |
           str_detect(alternate_category, "public_school"))

# inspect
poi_schools
```

```{r}
# identify duplicate postcodes
poi_schools <- poi_schools %>% 
  group_by(postcode) %>% 
  mutate(rank = rank(primary_name)) %>% 
  ungroup()

# filter school poi data
poi_schools <- poi_schools %>% 
  filter(is.na(postcode) | rank == 1) %>% 
  select(-rank)

# inspect
poi_schools
```

We can use a similar approach to approximate the locations of fast food outlets in the Borough

```{r}
# filter fast food poi data
poi_fastfood <- poi24 %>% 
  filter(str_detect(main_category, "fast_food_restaurant") | str_detect(alternate_category, "fast_food_restaurant") | str_detect(alternate_category, "chicken_restaurant") | str_detect(alternate_category, "burger_restaurant"))

# inspect
poi_fastfood
```

Map both datasets to get an idea of how the data looks like
```{r}
# combine for mapping
poi_schools <- poi_schools %>% 
  mutate(type = "School")
poi_fastfood <- poi_fastfood %>% 
  mutate(type = "Fast food")
poi_lambeth <- rbind(poi_schools, poi_fastfood)

tm_shape(lambeth) +
  tm_polygons(col = "#f0f0f0") +
tm_shape(poi_lambeth) +
  tm_dots(col = "type", size = 0.05, palette = c("#beaed4", "#fdc086"), title = "") +
tm_layout(legend.outside = TRUE,
          legend.position = c("right", "bottom"),
          frame = FALSE)
```

```{r}
# define our bbox coordinates, use WGS84
bbox_lambeth <- poi24 %>% 
  st_transform(4326) %>% 
  st_bbox()

# osm query
osm_network <- opq(bbox = bbox_lambeth) %>% 
  add_osm_feature(key = "highway", value = c("primary", "secondary", "tertiary", "residential", "path", "footway", "unclassified", "living_street", "pedestrian")) %>% 
  osmdata_sf()
```

```{r}
# extract the nodes, with their osm_id
osm_network_nodes <- osm_network$osm_points[, "osm_id"]

# extract the edges, with their osm_id and relevant columns
osm_network_edges <- osm_network$osm_lines[, c("osm_id", "name", "highway", "maxspeed", "oneway")]

# inspect
head(osm_network_nodes)
head(osm_network_edges)
```

We can quickly map the network edges to see how the road network looks like

```{r}
tm_shape(osm_network_edges) +
  tm_lines(
    col = "#bdbdbd",
    lwd = 0.2
  ) +
tm_shape(lambeth) +
  tm_borders(
    col = "#252525",
    lwd = 2
  ) +
tm_add_legend(
  type = "line",
  labels = "Road segments",
  col = "#bdbdbd"
) +
tm_add_legend(
  type = "line",
  labels = "Oueline Lambeth",
  col = "#252525"
) +
tm_layout(
  frame = FALSE,
  legend.outside = TRUE,
  legend.position = c("right", "bottom")
)

```

```{r}
# overwrite one-way default
osm_network_edges$oneway <- "no"
```

```{r}
# replace missing highway tags with unclassified
osm_network_edges <- osm_network_edges %>% 
  mutate(highway = if_else(is.na(highway), "unclasssified", highway))

# create network graph
osm_network_graph <- weight_streetnet(osm_network_edges, wt_profile = "foot")
```

Once we have constructed our graph, we can use it to calculate network distances between our points of interest

```{r}
# extract the largest connected graph component
netx_connected <- osm_network_graph[osm_network_graph$component == 1,]

# inspect number of remaining road segments
nrow(netx_connected)
```

Now that we have our connected subgraph, we can use the `dodgr_distance()` function to calculate the network distances between every possible origin (i.e. school) and destination (i.e. fast food outlet)

```{r}
# reproject as the dodgr package requires data to be projected in WGS84
poi_schools <- poi_schools %>% 
  st_transform(4326)
poi_fastfood <- poi_fastfood %>% 
  st_transform(4326)

# distance matrix
distance_matrix <- dodgr_distances(netx_connected, from = st_coordinates(poi_schools), to = st_coordinates(poi_fastfood), shortest = FALSE, pairwise = FALSE, quiet = FALSE)
```

The result of this computation is a distance matrix that contains the network distances between all origins (i.e. schools) and all destinations (i.e. fastfood outlets)

```{r}
# inspect
distance_matrix[1:5, 1:5]
```

```{r}
# fast-food outlets within 500m
poi_schools$fastfood_500m <- rowSums(distance_matrix <= 500)
# fast-food outlets within 1000m
poi_schools$fastfood_1000m <- rowSums(distance_matrix <= 1000)
```

In the final step, we can invetigate whether there is a relationship between the proximity of fast-food outlets and the relative levels of deprivation in the area

```{r}
# read imd dataset
imd19 <- read_csv(here::here("Dataset", "England-IMD-2019.csv"))

imd19
```

```{r}
# join imd
lsoa11 <- lsoa11 %>% 
  left_join(imd19, by = c("lsoa11cd" = "lsoa11cd"))

# join schools to their parent lsoa
poi_schools <- poi_schools %>% 
  st_transform(27700) %>% 
  st_join(lsoa11)
```

We can use this approach to derive the average number of fast-food by IMD decile

```{r}
# average counts by imd decile
fastfood_imd <- poi_schools %>% 
  group_by(imd_dec) %>% 
  mutate(avg_cnt = mean(fastfood_1000m)) %>% 
  distinct(imd_dec, avg_cnt) %>% 
  arrange(imd_dec)

# inspect
fastfood_imd
```

There appears to be a weak relationship, with schools in more deprived areas having, on average, a higher number of fast-food outlets within a 1,000-metre walking distance. However, this trend is not consistent, as schools in the least deprived areas of Lambeth show the highest accessibility on average.

